Quantity discrimination, decision-making, and the role of early-life conditions in a lizard

### 2025-09-24 ms.qmd
updates

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen ^3,4^, Amelia Y. Peardon ^1^, Daniel W.A. Noble^1^

@@

## Abstract
Quantity discrimination affects a range of behaviours essential for fitness, including social interactions, navigation, and foraging. Animals use different cognitive systems to distinguish between quantities. The Approximate Number System (ANS) estimates, while the Object File System (OFS) tracks discrete items. In reptiles, quantity discrimination remained largely unexplored until recently, and the use of one system over the other can depend on the species or the context in which they are tested. Early life conditions can also influence brain development and cognition, potentially affecting quantity discrimination and decision-making. Here, we tested the ability of the common garden skink (_Lampropholis guichenoti_) to discriminate between quantities and the effects of elevated prenatal GCs and incubation temperature on quantity discrimination and decision-making. We assessed quantity discrimination using a spontaneous choice test with food as a stimulus. We subjected lizards to five two-choice tests differing in the number of items (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4) while controlling for total length and area. Contrary to our predictions, we found no evidence of quantity discrimination in _L. guichenoti_. Furthermore, prenatal conditions did not influence their performance or decision-making. Despite these results, further investigation of quantity discrimination in live-prey-feeding reptiles and the impact of early conditions on other taxa remains an important area of research.

@@

The main two experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. Spontaneous choice tests ascertain if animals can distinguish between two groups of biologically relevant stimuli that differ quantitatively (number, size, volume, etc) [@agrillo2014spontaneous; @cox2016quantity]. Experiments employing training procedures evaluate an animal’s ability to learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other and can provide different results even in the same species [see @miletto2017quantitative; @miletto2018quantity].

Numerous reasons exist for why methods can lead to different results. First, spontaneous choice tests assume that one of the options presented is more suitable than the other (e.g. the group with more conspecifics or the larger amount of food) when this may not always be the case. For example, choosing groups of conspecifics with fewer individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [@yee2013costs]. As such, context and motivation can be essential drivers of decision-making that can change the way animals discriminate between quantities [@agrillo2014spontaneous]. Second, each method may engage distinct cognitive mechanisms when processing numerical information [@agrillo2014spontaneous]. Different procedures may shape how animals discriminate between quantities, making them more responsive towards certain quantity traits over others. Some tasks may encourage a focus on approximate magnitudes, while others may promote attention to individual items [@agrillo2014spontaneous].

In this regard, researchers propose that two core systems underlie numerical cognition: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. The ANS estimates the approximate magnitude of items within a set, with accuracy decreasing as the numerical difference between sets decreases — a principle known as Weber’s law [@hyde2011two]. In contrast, the OFS allows for the precise tracking of discrete items but is thought to be limited to small quantities, typically up to three or four [@hyde2011two]. The engagement of these systems depends not only on the numerical magnitude but also on contextual factors such as the availability of continuous variables (e.g., area, density, movement) and the nature of the task [@stancher2015numerical]. Such continuous cues typically favor ANS-based processing, while in their absence, the OFS is more likely to be employed [@agrillo2014spontaneous], which often complicates the study of numerosity, as continuous traits must be carefully controlled to ensure the use of the OFS [@agrillo2014spontaneous]. In spontaneous choice tests, for instance, animals may rely on cumulative surface area rather than numerical differences when selecting between two sets of food items [@agrillo2014spontaneous]. To disentangle these cues, researchers often keep the total area of the sets constant while changing the number of items, forcing individuals to rely on numerical information alone [@stancher2015numerical].

@@

Developmental conditions can strongly influence cognition because the brain is particularly sensitive to environmental inputs in early life stages [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density and learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. Thermal effects on cognition may arise directly or indirectly through parental effects [@Crino_2023]. For example, temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and developmental temperature may be relevant in shaping reptile cognition. Most studies on the effects of GC exposure or elevated temperatures during development focus on learning and memory, but few studies have investigated how early-life conditions can affect quantity discrimination despite its importance on animals' lives [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different numbers of food items. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different numbers of crickets even when controlling by other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Egg length and width were measured with digital calipers to the nearest mm (± 0.1 mm), while mass was recorded with a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Eggs were assigned a clutch number and an individual identity. Then, we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below) and placed them in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite). We covered the cups with plastic wrap to retain moisture and left the eggs in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week.

*Hatchlings* -- Immediately after hatching, we measured snout-vent length (SVL) and tail length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two developmental stress treatments: CORT-treated eggs were topically supplied with 5 µL of a CORT solution (10 pg/mL). Control treatment eggs received an equal volume of 100% Ethanol. Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. based on previous studies, we selected doses where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limits of the natural incubation temperatures for this species [@qualls2000post].

@@

Lizards were acclimatised to the experimental arenas for two weeks before starting the numerical discrimination task (see below). The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). Arenas were placed on seven different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms being the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Each numerical test was made only one per lizard. The food consisted of small thawed frozen crickets (*A. domestica*) that were dusted with calcium and multivitamins. We decided to use thawed frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell]. Lizards were habituated to eating thawed frozen crickets for one month before the experiments.

The crickets used during the experiment were all the same size, but to control for some continuous traits, we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see _Supplementary Material: Control of size in both options_). Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material: Prey orientation test_).

To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for side biases, the higher number of crickets was randomly placed on the right or the left platform. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day.

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. After the first interaction with any of the options, we recorded the time spent interacting with that option for a maximum of five minutes. We recorded this time for both amounts of food independently of the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time spent interacting with the smaller food option from the time spent interacting with the larger one. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. _Latency_, _Choice_, and _Interest_ were noted as NA when there was no interaction with any of the choices. We also considered _Interest_ as NA in a few cases where the animals managed to lift the dish and eat the prey; _Latency_ and _Choice_ were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

Motivation plays a key role in animals' performance in spontaneous choice tests [@agrillo2014spontaneous; @bisazza2014experim]. To ensure that lizards remained motivated to forage, they were provided only one cricket during acclimatisation and the numerical task (see above). During acclimatisation, lizards were fed at the same time each day as the experiments were conducted. In the numerical task, lizards received a cricket immediately after the test and then fasted for more than 20 hours, which we expected would help standardise fasting levels across individuals. To further control for motivation, we excluded from our analyses any trials in which lizards did not interact with any options or did not eat the cricket provided after the test (n = 0 in both cases). Consequently, we do not consider motivation a significant factor influencing our results (but see Discussion).

@@

We modelled the three main recorded behaviours: _Latency_, _Choice_, and _Interest_ as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)]. We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included sex, and age as predictors. We did not find any effect of sex or age on Choice or Interest, but we found an effect on Latency (see Results). Regardless, we included both factors in all models. Age was centered to zero in the model, and we averaged the predicted values for males and females.

We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, so clutch identity captures potential maternal effects.

@@

We recorded and analysed a total of 400 videos from 80 lizards (n = 20 per treatment, see @fig_results). The lizards were between `r min(data_num$age_raw)` and `r max(data_num$age_raw)` days post-hatching (see Fig. S5 in _Supplementary Material_). Final sample sizes were composed by 43 females and 37 males, with no less than 6 females per treatment and no more than 13.

@@

Neither temperature, CORT, or their interaction affected performance in any of the tests, nor were there any clear patterns suggestive of an effect of the developmental environment on numerical discrimination (see Table 1 and @fig-results). Except for Hot-CORT lizards in 1 VS 3 tests or Cold-Control lizards in 2 VS 3 tests, the predicted probability of choosing the larger number of crickets first was not different from 0.5 (@fig-results; but also see Table S1 and Table S2 in _Supplementary Material_). Similarly, interest in the larger number of crickets (_Interest_) was not different from 0 in any of the tests (@fig-results; but also see Table S1 and Table S2 in _Supplementary Material_).

@@

#### Foraging context may influence spontaneous quantity discrimination in _L. guichenoti_  
The lack of numerical discrimination in food-related spontaneous tests in _L. guichenoti_ may be due to their insectivorous diet. In studies where reptiles discriminated between different amounts of food, the relevant stimuli involved either vegetables [@gazzola2018continuous; @szabo_spontaneous_2021] or large numbers or alive prey (e.g., 5 vs. 10) [@recio_prey_2021]. In contrast, when _Podarcis siculus_ were presented with different numbers of _Musca domestica_ larvae in a spontaneous choice test, the lizards failed to discriminate between quantities despite showing quantity discrimination when prey size was the main cue [@miletto2017quantitative] or when tested using training procedures [@miletto2018quantity]. Notably, the numbers of food items employed in @miletto2017quantitative were always fewer than four. Spontaneous choice tests assume that animals should select the most appropriate quantity, which may not necessarily be the largest one [@agrillo2014spontaneous] if the costs of selecting the larger option outweigh the benefits. For example, if capturing larger prey demands more time or energy or success rates remain similar regardless of prey quantity, animals may not prefer larger quantities [@agrillo2014spontaneous]. Similarly, animals would not show any preference towards larger number of food items if the likelihood of acquiring one item is the same regardless of quantity [@agrillo2014spontaneous]. As a result, even if diet might not affect quantity discrimination abilities, insectivorous lizards might not have evolved a preference for larger patches of prey, at least with small numbers [see @recio_prey_2021].

>>>>

Pablo Recio^1, 2^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen ^3,4^, Amelia Y. Peardon ^1^, Daniel W.A. Noble^1^

@@

## Abstract
Quantity discrimination affects a range of behaviours essential for fitness, including social interactions, navigation, and foraging and is widespread across animal taxa. However, in reptiles, evidence for this ability—particularly for small number contrasts—remains limited and mixed. Early-life conditions, such as hormonal exposure and temperature, can also shape brain development and cognitive performance, potentially affecting numerical abilities. Here, we tested the ability of the common garden skink (_Lampropholis guichenoti_) to discriminate between quantities and the effects of elevated prenatal GCs and incubation temperature on quantity discrimination and decision-making. We assessed quantity discrimination using a spontaneous choice test with food as a stimulus. We subjected lizards to five two-choice tests differing in the number of items (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4) while controlling for total length and area. Contrary to our predictions, we found no evidence of quantity discrimination in _L. guichenoti_. Furthermore, prenatal conditions did not influence their performance or decision-making. Despite these results, further investigation of quantity discrimination in live-prey-feeding reptiles and the impact of early conditions on other taxa remains an important area of research.

@@

The main two experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. Spontaneous choice tests ascertain if animals can distinguish between two groups of biologically relevant stimuli that differ quantitatively (number, size, volume) [@agrillo2014spontaneous; @cox2016quantity]. Experiments employing training procedures evaluate an animal’s ability to learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other and can provide different results even in the same species [see @miletto2017quantitative; @miletto2018quantity].

Numerous reasons exist for why methods can lead to different results. First, spontaneous choice tests assume that one of the options presented is more suitable than the other (e.g. the group with more conspecifics or the larger amount of food) when this may not always be the case. For example, choosing groups of conspecifics with fewer individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [@yee2013costs]. As such, context and motivation can be essential drivers of decision-making that can change the way animals discriminate between quantities [@gazes2018impact]. Second, each method may engage distinct cognitive mechanisms when processing numerical information [@agrillo2014spontaneous]. Different procedures may shape how animals discriminate between quantities, making them more responsive towards certain quantity traits over others. Some tasks may encourage a focus on approximate magnitudes, while others may promote attention to individual items [@soldati2017long].

In this regard, researchers propose that two core systems underlie numerical cognition: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. The ANS estimates the approximate magnitude of items within a set, with accuracy decreasing as the numerical difference between sets decreases — a principle known as Weber’s law [@hyde2011two]. In contrast, the OFS allows for the precise tracking of discrete items but is thought to be limited to small quantities, typically up to three or four [@hyde2011two]. However, when discriminating between small quantities, the engagement of these systems also depends on contextual factors such as the availability of continuous variables (e.g., area, density, movement) and the nature of the task [@stancher2015numerical]. Such continuous cues typically favor ANS-based processing, while in their absence, the OFS is more likely to be employed [@agrillo2014spontaneous], which often complicates the study of numerosity, as continuous traits must be carefully controlled to ensure the use of the OFS [@agrillo2014spontaneous]. In spontaneous choice tests, for instance, animals may rely on cumulative surface area (ANS-related) rather than numerical (OFS-associated) differences when selecting between two sets of food items [@agrillo2014spontaneous]. To disentangle these cues, researchers often keep the total area of the sets constant while changing the number of items, forcing individuals to rely on numerical information alone [@vonk2012bears; @stancher2015numerical; @beran2016capuchin].

@@

Developmental conditions can strongly influence cognition because the brain is particularly sensitive to environmental inputs in early life stages [@zhu_prenatal_2004]. In ectotherms, the temperature at which embryos develop have significant effects on several traits [@noble_developmental_2018]. For example, incubation at warm temperatures influences neuron density and learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. Importantly, thermal effects on cognition may arise directly or through parental effects [@Crino_2023].Thermal fluctuations, particularly those approaching the edges of an organism's optimal range can trigger the stress response, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. For instance, stressing pregnant rats (_Rattus norvegicus_) suppresses neurogenesis in the dentate gyrus of the offspring, associated with impairments in spatial learning [@lemaire_prenatal_2000]. As such, the interaction between GCs and developmental temperature may be relevant in shaping reptile cognition, especially under the context of climate change [@Crino_2023]. Most studies on the effects of GC exposure or elevated temperatures during development focus on learning and memory, but few studies have investigated how early-life conditions can affect quantity discrimination despite its importance on animals' lives [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different numbers of food items. Our aim was to understand if this species could use the OFS to discriminate between small quantities of prey in a spontaneous choice test. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different numbers of crickets even when controlling for other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task as previous studies have reported enhanced cognitive abilities ectotherms incubated under warmer conditions [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Eggs were assigned a clutch number and an individual identity. Then, we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below)and placed each egg in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite) and covered the cups with plastic wrap to retain moisture. Each egg was incubated individually in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week. Egg and hatchling mortality was not higher than 13% in any of the treatments (Leibold et al., In prep.).

*Hatchlings* -- Hatchlings were placed in individual enclosures immediately after hatchling (18.7L x 13.2W x 6.3H cm) with non-stick matting and a small water dish and until the acclimatisation period (see below). During this time, they were given water daily and received 3-6 small _A. domestica_ crickets three times a week. All care otherwise followed similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two developmental stress treatments: CORT-treated eggs were topically supplied with 5 µL of a CORT solution (10 pg/mL) once, immediately after collection and before incubation. Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol [@carter2016turtle; @carter2018evidence]. This CORT dose increased mean yolk CORT levels 2 standard deviations higher than those of control eggs in previous studies [@crino2024eggs]. Control treatment eggs received an equal volume of 100% Ethanol. Then, all eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limits of the natural incubation temperatures for this species [@qualls2000post].

@@

Lizards were acclimatised to the experimental arenas for two weeks before starting the experiments (see below). The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). Arenas were placed on seven different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms being the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five numerical discrimination tests per lizard, each differing in the ratio between both choices of crickets: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Each numerical discrimination test was made only once per lizard. We counterbalanced the order of testing by ensuring that, each day, an equal number of individuals from each treatment were tested in each type of trial. We controlled for potential side biases by placing the higher number of crickets randomly on the right or the left platform. The food consisted of small thawed frozen crickets (_A. domestica_) that were dusted with calcium and multivitamins. We decided to use thawed frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell]. After acclimatisation to the arenas and before the numerical tests, lizards were habituated to eating thawed frozen crickets for one month before the experiments.

The crickets used during the experiment were all the same size, but to control for some continuous traits, we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see _Supplementary Material: Control of size in both options_). this analysis showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we subjected the same lizards to another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material: Prey orientation test_).

To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials.

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. Every trial began when we left the platform with the crickets in the enclosure, and lasted for 90 minutes until we took the platform away. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). Interaction was defined as touching the Petri dish with the snout or having the front half of the body on the platform. For each option, and regardless of _Latency_, we recorded the time the lizard spent interacting with it for five minutes after the first interaction. The timer was not paused if the lizard walked away, so the recorded time reflects total engagement with each option during the first five minutes following initial contact. We calculated _Interest_ by subtracting the time spent interacting with the smaller food option from the time spent interacting with the larger one. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. _Latency_, _Choice_, and _Interest_ were noted as NA when there was no interaction with any of the choices (n = 4 observations). We also considered _Interest_ as NA in a few cases where the animals managed to lift the dish and eat the prey (n = 7 observations); _Latency_ and _Choice_ were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests (n = 0 individuals excluded). All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

Motivation plays a key role in animals' performance in spontaneous choice tests [@bisazza2014experim; @gazes2018impact]. To ensure that lizards remained motivated to forage, they were provided only one cricket per day during acclimatisation and the numerical task (see above). During acclimatisation, lizards were fed at the same time each day as the experiments were conducted. In the numerical task, lizards received a cricket immediately after the test and then fasted for more than 20 hours, which we expected would help standardise fasting levels across individuals. To further control for motivation, we excluded from our analyses any trials in which lizards did not interact with any options or did not eat the cricket provided after the test (n = 0 in both cases). Consequently, we do not consider motivation a significant factor influencing our results (but see Discussion).

@@

We modelled the three main recorded behaviours: _Latency_, _Choice_, and _Interest_ as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)]. We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included sex, and age (days since hatching) as predictors. Age was centered to the mean to facilitate interpretation of model estimates, and posterior predictions were averaged across sexes to provide estimates independent of the reference category.

We included lizard identity and clutch identity as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, so clutch identity captures potential maternal effects.

@@

We recorded and analysed a total of 400 videos from 80 lizards (n = 20 per treatment, see @fig-results). The lizards were between `r min(data_num$age_raw)` and `r max(data_num$age_raw)` days post-hatching (see Fig. S5 in _Supplementary Material_). Final sample sizes were composed by 43 females and 37 males, with no less than 6 females per treatment and no more than 13.

@@

Neither temperature, CORT, or their interaction affected _Choice_ or _Interest_ in any of the tests, nor were there any clear patterns suggestive of an effect of the developmental environment on numerical discrimination (see Table 1 and @fig-results). Except for Hot-CORT lizards in 1 VS 3 tests or Cold-Control lizards in 2 VS 3 tests, the predicted probability of choosing the larger number of crickets first was not different from 0.5 (@fig-results; but also see Table S1 and Table S2 in _Supplementary Material_). Similarly, interest in the larger number of crickets (_Interest_) was not different from 0 in any of the tests (@fig-results; but also see Table S1 and Table S2 in _Supplementary Material_).

@@

#### Foraging context may influence spontaneous quantity discrimination in _L. guichenoti_  
We found no evidence of numerical discrimination in any of the tests performed. One potential concern is that pre-test fasting could have caused indiscriminate foraging masking numerical discrimination, but it seems unlikely. First, if hunger alone were driving behaviour, we would expect random first choices but clear preferences during subsequent interactions, reflected in the discrimination index. Second, only 30 and 36 out of 400 observations showed lizards interacting with the greater or fewer option, respectively, for more than 250 seconds (out of 300 s). Together, these results suggest that hunger levels were not masking numerical discrimination.
